Inferring the gravitational potential of the Milky Way with a few precisely measured stars

### 2014-02-13 text/paper2/ms.tex
some reorganizing

Cosmological simulations of galaxy formation in the $\Lambda$CDM paradigm predict dark matter halos that (1) are permeated with substructure on many scales, (2) are triaxial in shape, and (3) have density profiles and shapes that vary with radius \citep{dubinski91, jing02, kuhlen07, veraciro11}. Dark-matter-only simulations produce strongly triaxial halos \citep{} with large density variations \citep{zemp09}. Including baryons tends to soften the triaxiality and graininess in the inner galaxy through a combination of dissipative infall \citep{dubinski94} or cooling \citep{bryan13}. These processes combined with the gravity from a baryonic disk (ellipsoid) can act to make the inner halo more oblate (spherical), however these processes do not erase the clumpy, triaxial nature of the outer halo \citep[e.g.,][]{pontzen12}. This leads to radially-dependent axis ratios and orientation, density profile, and smoothness, thus suggesting complicated, non-intregrable potential forms for Milky Way-like galaxies.

@@

Within the Milky Way our unique vantage point allows us a three-dimensional view of the structure in own dark matter halo. Our proximity allows us to use individual stars as kinematic tracers and hence build much larger samples that probe deeper into the halo than the globular cluster and planetary nebula studies of external galaxies. For example, \cite{deason12a} used halo BHB stars selected from the Sloan Digital Sky Survey \cite[SDSS;][]{york00} as a random tracer population to measure the mass and slope of a simple, power-law fit to the Galactic potential. Such studies assume that the tracer orbits are randomly sampled from a smooth distribution function and phase mixed, however the existence of large-scale substructure in stellar streams and clouds is known to bias mass and velocity inferences that ignore this fact. Another approach then is to exploit the non-random nature of the stellar populations in the halo. Tidal streams are dynamically cold systems --- debris typically have small distributions of energy and angular momentum --- and thus require orders of magnitude fewer tracers than a random sample to get constraints of comparable accuracy. For example, in the simplest case we might {\it assume} that debris stars are actually still on the same orbit as their progenitor system \citep[a \emph{wrong} assumption, see e.g.][]{eyre}. Then we can imagine measuring the full-space velocities ${\bf v}$ at different points ${\bf x}$ in the structure (e.g. along a stream), which would give us a direct measure of differences in a potential, $\Phi$. In practice, simple methods such as orbit fitting lead to systematic biases in inferred properties of the underlying potential because streams do not strictly delineate orbits [cite eyre binney work].

Since the initial discovery of tidal debris from Sgr, myriad other streams and debris systems have been discovered in large photometric surveys probing Galactocentric distances from $\sim$15-100 kpc. The known debris structures are $\gtrsim$10 kpc from the Sun, thus while overdensities in the halo are less contaminated by background, the constituent stars often have incomplete or poorly measured phase-space information. Perhaps the most rigorous method of modeling streams and the potential is to run full N-body simulations of satellite disruption and compare the density of observed stars to simulated stars. Such a model could naturally incorporate observational uncertainties, missing dimensions, and would, in principle, allow for studying any arbitrary time-dependent or non-integrable potential. However, such a model is presently computationally intractable [though see milkyway@home stuff...]. \cite{law10} ran several N-body models of the disruption of the Sagittarius (Sgr) dwarf spheroidal galaxy and qualitatively matched their simulated streams to all Sgr data available at that time. By comparing the phase-space positions of their stream particles with the data, they were able to use these simulations to constrain the shape of the Milky Way potential out to $\sim$70 kpc [check number] and found that the best-fitting halo has a peculiar triaxial shape elongated in the Galactic Z direction, orthogonal to the plane of the disk. (though, see ... for possible complications in the Sgr debris belokurov 2012?).

The difficulty and computational costs associated with proper N-body modeling has motivated numerous efforts for developing approximate methods for constraining properties of the Galactic potential with cold debris structures. The simplest approximation is to fit a single orbit to observed debris \citep[e.g.,][]{koposov10, deg13}. This has the advantage of naturally incorporating observational errors, but is known to be invalid as debris does not exactly follow an orbit \citep[e.g.,][]{binney08, eyre11, sanders13a}. To account for this, methods have been proposed that model the debris as an orbit plus some dispersion or offset in phase space \citep[e.g.,][]{eyre09a, varghese11, kuepper12} or action-angle coordinates \citep{eyre11, sanders13b, bovy14}. Other statistical methods have been proposed \citep[][]{penarrubia12, sanderson14} that may prove powerful when applied to, e.g., data from the \gaia mission, where full 6D coordinates will be known for large samples of stars in the halo --- and therefore many debris structures --- but membership is not necessarily known. However, these methods do not account for observational errors.

Methods that rely on action-angle coordinates, [...]. Any generative model that relies on matching densities --- whether in phase-space, action-space, or observed coordinates --- must also include details about the progenitor and a model for the background of stars. [...] In \citet[][hereafter \paperone]{apw13}, we introduced a simple method for using individual stars combined with knowledge about the mass and orbit of the progenitor to constrain properties of the host galaxy potential (\emph{Rewinder}). [...]

- Then a short note on worries\\
\indent\indent- for generative models\\
\indent\indent- for any methods based on action/angle \\
\indent\indent- GD-1 was fit as if it highlighted an orbit.  That is known to be wrong.\\
\indent\indent- Densities are hard to match: need a background model, way to go from luminosity to density, needs assumptions about the mass function and structure of the satellite\\
\indent\indent- Triaxiality / radially-dependent shapes and substructure present problems for integral-based methods\\
- Then how rewinder gets round these????\\
\indent\indent- Reference Deg \& Widrow to say they illustrate the idea of the power of multiple streams to back up the bit where you say that's the long-term aim\\

[Another prediction of hierarchical structure formation in $\Lambda$CDM is the presence of substructure. ]

%[Chemical tagging will always only give probabilistic information.
%  But what if it delivers somewhat confident information about small
%  stellar families in the halo?]

@@

@@

\subsection{Motivation}
In \cite{apw13}, we introduced the \emph{Rewinder} algorithm [...]. In that work, we define a phase-space distance, $D_{ps}$, as the sum of the normalized, relative position, $\frac{\bsr-\bsr_p}{\rtide}$, and velocity, $\frac{\bsv-\bsv_p}{v_{\rm esc}}$, for a stream star. We minimized the generalized variance of the 6D distribution over each normalized coordinate evaluated at the time that each star has minimum $D_ps$. This simple idea proved quite powerful for constraining properties of the LM10 potential used simulated observations of their N-body simulation. However, this method does not properly incorporate observational uncertainties.

@@

\begin{figure}[h]
\begin{center}
%\includegraphics[width=0.45\textwidth]{fig4.eps}
\includegraphics[width=\textwidth]{/Users/adrian/projects/streams/plots/paper2/rel_dists.pdf}
\caption{ TODO }\label{fig:reldist}
\end{center}
\end{figure}

@@

\begin{figure}[h]
\begin{center}
%\includegraphics[width=0.45\textwidth]{fig4.eps}
\includegraphics[width=\textwidth]{/Users/adrian/projects/streams/plots/paper2/potentials.pdf}
\caption{ TODO }\label{fig:potential}
\end{center}
\end{figure}

@@

>>>>

Cosmological simulations of galaxy formation in the $\Lambda$CDM paradigm predict dark matter halos that (1) are permeated with substructure on many scales, (2) are triaxial in shape, and (3) have density profiles and shapes that vary with radius \citep{dubinski91, jing02, kuhlen07, veraciro11}. Dark-matter-only simulations produce triaxial halos \citep{jing02} with large density fluctuations \citep{zemp09}. Including baryons tends to soften the triaxiality and graininess in the inner galaxy through a combination of dissipative infall \citep{dubinski94} or cooling \citep{bryan13}. These processes combined with the gravity from a baryonic disk (ellipsoid) can act to make the inner halo more oblate (spherical), however these processes do not erase the clumpy, triaxial nature of the outer halo \citep[e.g.,][]{pontzen12}. This leads to radially-dependent axis ratios and orientation, density profile, and smoothness, thus suggesting complicated, non-intregrable potential forms for Milky Way-like galaxies.

@@

Within the Milky Way our unique vantage point allows us a three-dimensional view of the structure in own dark matter halo. Our proximity allows us to use individual stars as kinematic tracers and hence build much larger samples that probe deeper into the halo than the globular cluster and planetary nebula studies of external galaxies. For example, \cite{deason12a} used halo BHB stars selected from the Sloan Digital Sky Survey \cite[SDSS;][]{york00} as a random tracer population to measure the mass and slope of a simple, power-law fit to the Galactic potential. Such studies assume that the tracer orbits are randomly sampled from a smooth distribution function and phase mixed, however the existence of large-scale substructure in stellar streams and clouds is known to bias mass and velocity inferences that ignore this fact. Another approach is to exploit the non-random nature of the stellar populations in the halo. Tidal streams are dynamically cold systems --- debris typically have small distributions of energy and angular momentum --- and thus require orders of magnitude fewer tracers than a random sample to get constraints of comparable accuracy. For example, in the simplest case we might assume that debris stars are actually still on the same orbit as their progenitor system \citep[a \emph{wrong} assumption, see e.g.,][]{}. Then we can imagine measuring the full-space velocities ${\bf v}$ at different points ${\bf x}$ in the structure (e.g., along a stream), which would give us a direct measure of differences in a potential, $\Phi$. 
%In practice, simple methods such as orbit fitting lead to systematic biases in inferred properties of the underlying potential because streams do not strictly delineate orbits [cite eyre binney work].

Since the initial discovery of tidal debris from Sgr, myriad other streams and debris systems have been identified in large photometric surveys, probing Galactocentric distances from $\sim$15-100 kpc. The known debris structures are $\gtrsim$10 kpc from the Sun, thus while overdensities in the halo are less contaminated by background, the constituent stars often have incomplete or poorly measured phase-space information. Perhaps the most rigorous method of modeling streams and the potential is to run full N-body simulations of satellite disruption and compare the density of observed stars to simulated stars in whatever coordinates are measured. A generative N-body model could naturally incorporate observational uncertainties, missing dimensions, and would, in principle, allow for studying any arbitrary time-dependent or non-integrable potential. However, such a model is presently computationally intractable. \cite{law10} demonstrated the power of this idea by running a grid of N-body models of the disruption of the Sagittarius (Sgr) dwarf spheroidal galaxy and matched their simulated streams to all Sgr data available at that time. By comparing the phase-space positions of their stream particles with the data, they were able to constrain the shape of the Milky Way potential out to $\sim$70 kpc and found that the best-fitting halo has a prolate, triaxial shape elongated in the Galactic Z direction, orthogonal to the plane of the disk. 
% (though, see ... for possible complications in the Sgr debris belokurov 2012?).

The difficulty and computational costs associated with proper N-body modeling has motivated numerous efforts for developing approximate methods for constraining properties of the Galactic potential with cold debris structures. The simplest approximation is to fit a single orbit to observed debris \citep[e.g.,][]{koposov10, deg13}. This has the advantage of naturally incorporating observational errors, but is known to be invalid as debris does not exactly follow an orbit \citep[e.g.,][]{eyre11, sanders13a}. To account for this, methods have been proposed that model the debris as an orbit plus some dispersion or offset, either in phase space \citep[e.g.,][]{eyre09a, varghese11, kuepper12} or action-angle coordinates \citep{eyre11, sanders13b, bovy14}. Other statistical methods have been proposed \citep[][]{penarrubia12, sanderson14} that may prove powerful when applied to, e.g., data from the \gaia mission, where full 6D coordinates will be known for large samples of stars in the halo --- and therefore many debris structures --- but stream membership is not known for all stars (however these statistical methods do not account for observational errors).

[Here we need to talk about integrable potentials and shortcomings of the methods above?] Methods that rely on action-angle coordinates, [...]. Any generative model that relies on matching densities --- whether in phase-space, action-space, or observed coordinates --- must also include details about the progenitor and a model for the background of stars. [...] 
- Then a short note on worries\\
\indent\indent- for generative models\\
\indent\indent- for any methods based on action/angle \\
\indent\indent- GD-1 was fit as if it highlighted an orbit.  That is known to be wrong.\\
\indent\indent- Densities are hard to match: need a background model, way to go from luminosity to density, needs assumptions about the mass function and structure of the satellite\\
\indent\indent- Triaxiality / radially-dependent shapes and substructure present problems for integral-based methods\\
- Then how rewinder gets round these????\\
\indent\indent- Reference Deg \& Widrow to say they illustrate the idea of the power of multiple streams to back up the bit where you say that's the long-term aim

@@

In \citet[][hereafter \paperone]{apw13}, we introduced a simple method for using individual stars associated with tidal debris combined with knowledge about the mass and orbit of the progenitor to constrain properties of the host galaxy potential (\emph{Rewinder}). [...]

@@

\subsection{Motivation}

In \paperone, we define a phase-space distance, $D_{ps}$, as the sum of the normalized, relative position, $\frac{\bsr-\bsr_p}{\rtide}$, and velocity, $\frac{\bsv-\bsv_p}{v_{\rm esc}}$, for a stream star. We minimized the generalized variance of the 6D distribution over each normalized coordinate evaluated at the time that each star has minimum $D_ps$. This simple idea proved quite powerful for constraining properties of the LM10 potential used simulated observations of their N-body simulation. However, this method does not properly incorporate observational uncertainties.

@@

\begin{figure}[h]
\begin{center}
% \includegraphics[width=\textwidth]{/Users/adrian/projects/streams/plots/paper2/rel_dists.pdf}
\caption{ TODO }\label{fig:reldist}
\end{center}
\end{figure}

@@

\begin{figure}[h]
\begin{center}
% \includegraphics[width=\textwidth]{/Users/adrian/projects/streams/plots/paper2/potentials.pdf}
\caption{ TODO }\label{fig:potential}
\end{center}
\end{figure}

@@

[Another prediction of hierarchical structure formation in $\Lambda$CDM is the presence of substructure. In principle, this method would work on any potential that we can integrate orbits in.]
